The Second Data Release of the INT Photometric H$\alpha$ Survey of the Northern Galactic Plane (IPHAS DR2)

### 2014-03-13 paper/iphas-dr2.tex
Update iphas-dr2.tex

\author[G. Barentsen
et. al]{Geert Barentsen$^{1}$\thanks{E-mail:geert@barentsen.be},
H. J. Farnhill$^1$,
J. E. Drew$^1$,
E. A. Gonz$\acute{\rm{a}}$lez-Solares$^2$, \newauthor
R. Greimel$^3$,
M. J. Irwin$^2$,
B. Mizalski$^{99}$,
C. Ruhland$^1$,
P. Groot$^4$,
A. Mampaso$^5$, \newauthor
S. E. Sale$^6$,
plus many contributors.
\newauthor\\
$^{1}$School of Physics, Astronomy \& Mathematics, University of Hertfordshire, College Lane, Hatfield, Hertfordshire, AL10 9AB, U.K.\\
$^{2}$Institute of Astronomy, Cambridge University, Madingley Road, Cambridge, CB3 OHA, U.K.\\
$^{3}$IGAM, Institute of Physics, University of Graz, Universit\"atsplatz 5, Graz, Austria.\\
$^{4}$Afdeling Sterrenkunde, Radboud Universiteit Nijmegen, Faculteit NWI, Postbus 9010, 6500 GL Nijmegen, The Netherlands.\\
$^{5}$Instituto de Astrof\'isica de Canarias, 38200 La Laguna, Tenerife, Spain.\\
$^{6}$Rudolf Peierls Centre for Theoretical Physics, Keble Road, Oxford, OX1 3NP, U.K.\\
$^{99}$Please add your current affiliation.\\
}

@@

\begin{abstract} % BACKGROUND, OBJECTIVE, METHODS, RESULTS, CONCLUSIONS
The INT/WFC Photometric H$\alpha$ Survey 
of the Northern Galactic Plane (IPHAS)
is a 1800 square degrees imaging survey
covering the entire northern Milky Way at $|b| < 5^\circ$
in the $r$, $i$ and \ha\ filters 
using the Wide Field Camera (WFC) 
on the 2.5-meter Isaac Newton Telescope (INT).
This data release presents the first 
uniformly-calibrated source catalogue
to have been extracted from the images,
providing single-epoch photometry
for 219 million unique sources
across 92\% of the survey.
The observations were carried out between 2003 and 2012
at a median seeing of 1.1 arcsec
to a depth of $r=21.2\pm 0.5$, $i=20.0\pm 0.3$ and \ha$=20.3\pm 0.3$
($5\sigma$ limits, Vega system).
We explain the data reduction 
and quality control procedures,
describe and test the new uniform photometric calibration,
and detail the construction of the source catalogue
and its quality warning flags.
We find that the new calibration
and the photometric repeatability are accurate to
$\sigma=0.03$~mag.
Finally, we demonstrate the ability of the 
catalogue's unique
$r$-$i$/$r$-\ha\ colour-colour diagram to
(i) characterise stellar populations and extinction regimes
towards different Galactic sightlines
and (ii) select candidate \ha-emission line objects.
The catalogue which accompanies this paper
provides the much-needed visible-light counterpart
to several infrared surveys of the Galactic Plane,
including many poorly-studied star-forming regions and ``bubbles''.
It also provides images for some of the most
crowded regions to be faced by Gaia.
\end{abstract}

@@

Since the first data release in 2008, 
the INT/WFC Photometric H$\alpha$ Survey 
of the Northern Galactic Plane (IPHAS)
has provided new insights into the contents and structure 
of our own backyard, the Milky Way. 
The original motivation for undertaking 
this large-scale programme of observation
-- spanning almost a decade 
and using more than 300 nights 
at the Isaac Newton Telescope (INT) in La Palma -- 
was to provide the digital update 
to the photographic northern H$\alpha$ surveys 
of the mid-20th century. 
By increasing the sensitivity 
with respect to these previous surveys 
by a factor $\sim$1000 (7 magnitudes), 
it was envisaged that IPHAS would allow 
the limited bright samples of Galactic emission line objects 
available at the outset \citep[e.g.][]{Kohoutek1999}, 
to be extended into larger, deeper, more statistically-robust samples 
that in turn could better inform our understanding 
of the early and late stages of stellar evolution. 
This aim has begun to be realised through a
range of published studies including: 
a preliminary catalogue of candidate emission line objects \citep{Witham2008};
discoveries of new northern symbiotic stars \citep{Corradi2008, Corradi2010}; 
new cataclysmic variables \citep{Witham2007}; 
new groups of young stellar objects \citep{Vink2008,Barentsen2011a};
along with discoveries of new and remarkable planetary nebulae 
\citep{Mampaso2006, Corradi2011, Viironen2011}.

Over the years it has become apparent that the legacy of IPHAS 
will extend beyond the traditional \ha\ applications 
of identifying emission line stars and nebulae. 
Through the provision of $r$, $i$ broadband photometry 
alongside H$\alpha$ data,
IPHAS has created the opportunity 
to study Galactic Plane populations 
in a new way.
For example, the survey’s unique $r-H\alpha$ colour, 
when combined with $r-i$,
has been shown to provide simultaneous constraints 
on intrinsic stellar colour and interstellar extinction \citep{Drew2008}. 
This has opened the door 
to a wide range of Galactic science applications, 
including the mapping of extinction across the Plane in three dimensions
and the probabilistic inference of stellar properties
\citep{Sale2009, Sale2010, Giammanco2011, Sale2012, Barentsen2013}. 
In effect, the availability of narrowband H$\alpha$ 
alongside $r, i$ magnitudes
provides coarse spectral information for huge samples of stars 
which are otherwise too faint or numerous 
to be targeted by spectroscopic surveys.
For such science applications to succeed however, 
it is vital that the imaging data is transformed 
into a homogeneously calibrated photometric catalogue, 
in which quality problems 
and duplicate detections of the same source 
are flagged.

The first release of IPHAS data, 
covering roughly half the survey footprint,
was made in late 2007 \citep{Gonzalez-Solares2008}. 
At the time the data were insufficiently complete 
to support a homogeneously calibrated source catalogue.
The goal of this paper is to present the next release 
that takes the coverage up to over 90 percent of the survey area 
and includes a uniform calibration.
In this work we aim to
(i) explain the data reduction 
and quality control procedures that were applied,
(ii) describe and test the new global photometric calibration, and 
(iii) detail the construction and demonstrate the use 
of the source catalogue that has been extracted 
from the newly re-calibrated data.

In \S\ref{sec:observations} we start by recapitulating the key points
of the survey design and strategy.
In \S\ref{sec:reduction} we describe the data reduction
and quality control procedures.
In \S\ref{sec:calibration} we explain the uniform re-calibration
and test our results against the Sloan Digital Sky Survey (SDSS).
In \S\ref{sec:catalogue} we explain how the source catalogue was compiled.
In \S\ref{sec:discussion} we discuss the properties of the catalogue,
and finally in \S\ref{sec:demonstration} we demonstrate
the health of the release by demonstrating its scientific exploitation.
In \S\ref{sec:conclusions} we conclude and outline
our future ambitions.

@@

The Wide Field Camera is a mosaic of 4 CCDs 
that captures a sky area of close to 0.29 square degrees.
To cover the entire northern Plane with some overlap,
the survey area was divided into 7635 telescope pointings.
Each of these pointings is accompanied by an offset position
at a displacement of $+$5 arcmin in declination 
and $+$5 arcmin in right ascension,
to deal with inter-CCD gaps, detector imperfections,
and to enable quality checks. 
The basic unit of observation hence
amounts to $2 \times 3$ exposures, 
in which each of the 3 survey filters is exposed at 2 offset sky positions, 
within an elapsed time of 10 minutes.
We shall refer to the unit of 3 exposures at the same position 
as a \emph{field},
and the combination of two fields at a small offset as a \emph{field pair}.
The survey hence contains 15270 fields grouped into 7635 field pairs.
To achieve the desired survey depth of 20th magnitude or fainter, 
the filter exposure times were set at 120 sec (narrowband \ha), 
30 sec ($r$) and 10 sec ($i$)
in the majority of the survey observations.\footnote{In 2003 
the $r$-band exposure time was 10~sec instead of 30~sec,
and since Oct 2010 the $i$-band exposure time 
has been increased from 10~sec to 20~sec.}

Data-taking began in the second half of 2003, 
and every field had been observed at least once by the end of 2008.
At that time only 76 percent of the field pairs 
satisfied our minimum quality criteria however,  
often due to the effects of clouds, poor seeing, or technical faults
(the quality criteria will be detailed in the next section). 
Since then, a programme of repeat observations has been in place 
to improve data quality. 
As a result, 92\% of the survey 
now benefits from quality-approved data.
The most recent observations which are part of this release
were obtained in November 2012.

\begin{figure*}
        \includegraphics[width=1\linewidth]{./plots/footprint/footprint_small.png}
        \caption{Survey area showing the footprints
        of all the quality-approved IPHAS fields
        which have been included in this data release.
        The area covered by each field has been coloured black
        with a semi-transparent opacity of 20\%,
        such that regions where fields overlap are darker.
        The IPHAS strategy is to observe each field twice
        with a small offset,
        and hence the vast majority of the area 
        is covered twice (dominant gray colour).
        There are small overlaps between all the neighbouring fields
        which can be seen as a honeycomb-style pattern
        of dark gray lines across the survey area.
        Regions with incomplete data are apparent as white gaps (no data) 
        or as the lightest shade of gray
        (denoting that only the offset position is missing).}
        \label{fig:footprint}
\end{figure*}

Figure~\ref{fig:footprint} shows the footprint 
of the quality-approved observations included in this work. 
The fields which remain missing 
-- covering 7 percent of the survey area --
are predominantly located towards the Galactic anti-center 
at $l > 120^o$.
Fields at these longitudes can only be accessed from La Palma 
in the months of November-December,
which is when the weather and seeing conditions are often poor
at the INT and observing attempts have failed repeatedly.
To enable the survey to be brought to completion, 
a decision was made recently to limit repeats in this area 
to individual fields requiring replacement,
i.e. fresh observations in all 3 filters may only be obtained 
at one of the two offset positions, 
if the data for the partner offset has already passed quality control.  
The catalogue is structured such that it is clear 
where a contemporaneous observation of both halves of a field pair
is not available.

@@

The reduced images are then stored in a multi-extension FITS file 
with a primary header describing the characteristics
(position, filter, exposure time, etc.) 
and four compressed image extensions 
corresponding to each of the four CCDs.
Source detection and characterisation is then carried out 
using the \textsc{imcore} tool \citep{Irwin1985,Irwin1997}.
The flux of each source is measured using both
the peak pixel height (i.e. a square 0.33$\times$0.33\arcsec\ aperture)
as well as a series of circular apertures of increasing diameter 
(1.2\arcsec, 2.3\arcsec, 3.3\arcsec, 4.6\arcsec\ and 6.6\arcsec).

The local background levels are estimated 
by computing the sigma-clipped median
flux in a grid of 64$\times$64 pixels (21$\times$21\arcsec)
across the image,
which is then interpolated to obtain an estimate 
of the background level at each pixel.
These sky levels are subtracted from the aperture photometry and
-- when required --
a deblending routine is applied which also attempts to remove
the contamination from any other nearby sources.
Whilst this approach works very well 
across the vast majority of the survey area,
the Galactic Plane unavoidably contains crowded regions 
with large numbers of overlapping sources
or rapidly spatially-varying nebulosity,
in which case aperture photometry must always be interpreted 
with caution.
In \S\ref{sec:catalogue} we will explain that overlapping
sources to which the deblending routine was applied 
are flagged in the catalogue using the \emph{deblend} warning flag.

@@

At the time of preparing DR2,
the CASU pipeline had processed
74\,195 IPHAS exposures 
in which a total of 1.9~billion \emph{candidate sources} were detected 
at the sensitive default detection level of 1.25\,$\sigma$
-- unavoidably including spurious detections, artefacts and
duplicate detections 
(in \S\ref{sec:catalogue} we will explain
how these have been removed or flagged).
The pipelined data set -- comprising 2.5~terabyte of FITS files --
was then transferred to the University of Hertfordshire
for the purpose of transforming the raw
detection tables into a reliable source catalogue which is 
(i) quality-controlled,
(ii) homogeneously calibrated, and 
(iii) contains user-friendly columns and warning flags.
It is these post-processing steps which are explained below.

@@

Observing time for IPHAS was obtained
on a semester-by-semester basis
through the traditional time allocation committees 
of the Isaac Newton Group of telescopes,
which are competitive and invariably over-subscribed.
For this reason, we attempted to utilise 
\emph{all} the nights allocated to IPHAS,
even those which were partially or entirely non-photometric
or otherwise affected by technical problems 
(e.g. electronic noise or telescope tracking problems).
Any unsuitable data that was taken as a result
of this strategy was subsequently flagged and rejected
using a series of seven quality criteria,
which ensure a reliable and homogeneous level of quality
across the data release:

@@

(1) \emph{Depth.} 
We discarded any exposures for which the $5\sigma$ limiting magnitude 
was worse than 20th magnitude in the $r$-band
or worse than 19th in $i$ or \ha. 
Such data were typically obtained during poor weather or full moon.
Most observations fared significantly better than these limits.
Figure~\ref{fig:depth} presents the distribution of limiting magnitudes
for all quality-approved fields,
which shows a mean depth of 
$21.2\pm0.5$ ($r$), $20.0\pm0.3$ ($i$) and $20.3\pm0.3$ (\ha).
We found that the depth achieved depended 
most strongly on the presence of the moon,
which was above the horizon during 62 percent 
of our observations and explains the wide and bi-modal shape
of the $r$-band limiting magnitude distribution 
(top panel in Fig.~\ref{fig:depth}).
In contrast, the depth attained in $i$ and \ha\ 
is less sensitive to moonlight
and the distribution of their depths
is hence more narrow
(middle and bottom panel in Fig.~\ref{fig:depth}).

Our minimum limiting magnitude criteria
have led us to exclude 9\% of the pipelined data.
We note that much of the excluded data may nevertheless be useful
for e.g. time-domain studies of bright stars.
The detection tables for any such discarded data are made
available through our website (www.iphas.org)
but are ignored in what follows.

(2) \emph{Ellipticity.} 
The ellipticity of a point source,
defined as $e = 1 - b / a$ 
with $b$ the semi-minor and $a$ the semi-major axis,
is a morphological measure of the elongation of the point spread function.
It is expected to be zero (circular) across the field 
in a perfect imaging system,
but it is slightly non-zero in any real telescope data 
due to optical distortions and tracking errors.
The mean ellipticity across a field in the IPHAS data set 
is $0.09\pm0.04$.
There have been sporadic episodes with higher ellipticities however
due to mechanical glitches in the telescope tracking system.
For this reason, we rejected exposures in which the mean ellipticity
across the detector exceeded $e > 0.3$,
which is when the photometric measurements delivered by the pipeline
were found to become degraded.
Only 0.4\% of the exposures were discarded on this basis.

(3) \emph{Seeing.} 
The survey originally aimed to obtain data 
at seeing better than 1.7 arcsec.
This target is currently attained across 86\% of the footprint,
in particular at early longitudes,
e.g. 92\% of the fields at $l<120^\circ$ are better than 1.7 arcsec.
Figure~\ref{fig:seeing} presents the distribution
of the mean seeing for all the quality-approved fields.
We find a median value of 1.1~arcsec in $r$/\ha\
and 1.0~arcsec in $i$.
In the $r$-band, 90\% of the data is better than 1.5~arcsec
and 10\% is better than 0.8~arcsec.
To improve the area covered by this data release,
we have decided to include the small fraction of data
that was obtained under seeing up to 2.5 arcsec,
so that only 1 per cent of the pipelined exposures
had to be excluded.
In \S\ref{sec:catalogue} we will explain
that the catalogue is structured
such that the information on the seeing is included,
and that the photometry listed
is based on the exposures with the best-available seeing.

>>>>

\author[G. Barentsen
et. al]{Geert Barentsen$^{1}$\thanks{E-mail:geert@barentsen.be},
H. J. Farnhill$^1$,
J. E. Drew$^1$,
E. A. Gonz$\acute{\rm{a}}$lez-Solares$^2$, \newauthor
R. Greimel$^3$,
M. J. Irwin$^2$,
B. Mizalski$^{99}$,
C. Ruhland$^1$,
P. Groot$^4$,
A. Mampaso$^5$, \newauthor
S. E. Sale$^6$,
M. J. Barlow,
R. L. M Corradi,
J. J. Drake,
J. Fabregat,
B. T. Gaensicke,
A. S. Hales,
J. Irwin,
C. Knigge,
D. J. Lennon,
J. R. Lewis,
M. Mohr-Smith,
R. A. H. Morris,
T. Naylor,
Q. A. Parker,
S. Phillipps,
R. Raddi,
P. Rodriguez-Gil,
L. Sabin,
D. Steeghs,
Y. C. Unruh,
K. Viironen,
J. S. Vink,
N. A. Walton,
N. J. Wright,
A. A. Zijlstra,
plus all those we forgot.
\newauthor\\
$^{1}$School of Physics, Astronomy \& Mathematics, University of Hertfordshire, College Lane, Hatfield, Hertfordshire, AL10 9AB, U.K.\\
$^{2}$Institute of Astronomy, Cambridge University, Madingley Road, Cambridge, CB3 OHA, U.K.\\
$^{3}$IGAM, Institute of Physics, University of Graz, Universit\"atsplatz 5, Graz, Austria.\\
$^{4}$Afdeling Sterrenkunde, Radboud Universiteit Nijmegen, Faculteit NWI, Postbus 9010, 6500 GL Nijmegen, The Netherlands.\\
$^{5}$Instituto de Astrof\'isica de Canarias, 38200 La Laguna, Tenerife, Spain.\\
$^{6}$Rudolf Peierls Centre for Theoretical Physics, Keble Road, Oxford, OX1 3NP, U.K.\\
$^{99}$Please add your current affiliation.\\
}

@@

\begin{abstract} % BACKGROUND, OBJECTIVE, METHODS, RESULTS, CONCLUSIONS
The INT/WFC Photometric H$\alpha$ Survey 
of the Northern Galactic Plane (IPHAS)
is a 1800 square degrees imaging survey
covering the entire northern Milky Way at $|b| < 5^\circ$
in the $r$, $i$ and \ha\ filters 
using the Wide Field Camera (WFC) 
on the 2.5-meter Isaac Newton Telescope (INT).
It is the first survey to offer comprehensive CCD photometry
of point sources across the Galactic Plane at visible wavelengths.
We present the first quality-controlled and
uniformly-calibrated source catalogue
to have been extracted from the images,
providing single-epoch photometry
for 219 million unique sources
across 92\% of the survey area.
The observations were carried out between 2003 and 2013
at a median seeing of 1.1 arcsec
to a depth of $r=21.2\pm 0.5$, $i=20.0\pm 0.3$ and \ha$=20.3\pm 0.3$
($5\sigma$ limits, Vega system).
We explain the data reduction 
and quality control procedures,
describe and test the new uniform calibration,
and detail the construction of the source catalogue
and its quality warning flags.
We find that the calibration
and the photometric repeatability are accurate to
$\sigma=0.03$~mag.
Finally, we demonstrate the ability of the 
catalogue's unique
$r$-$i$/$r$-\ha\ colour-colour diagram to
(i) characterise stellar populations and extinction regimes
towards different Galactic sightlines
and (ii) select candidate \ha\ emission-line objects.
The catalogue which accompanies this paper
provides the much-needed visible-light counterpart
to several infrared surveys of the Galactic Plane,
and provides images for some of the most
crowded regions to be faced by Gaia.
\end{abstract}

@@

Since the first data release in 2008, 
the INT/WFC Photometric H$\alpha$ Survey 
of the Northern Galactic Plane (IPHAS)
has provided new insights into the contents and structure 
of our own backyard, the Milky Way. 
The original motivation for undertaking 
this large-scale programme of observation
-- spanning almost a decade 
and using more than 300 nights 
at the Isaac Newton Telescope (INT) in La Palma -- 
was to provide the digital update 
to the photographic northern H$\alpha$ surveys 
of the mid-20th century. 
By increasing the sensitivity 
with respect to these previous surveys 
by a factor $\sim$1000 (7 magnitudes), 
it was envisaged that IPHAS would allow 
the limited bright samples of Galactic emission line objects 
available at the outset \citep[e.g.][]{Kohoutek1999}, 
to be extended into larger, deeper, more statistically-robust samples.
This in turn could better inform our understanding 
of the early and late stages of stellar evolution. 
This aim has begun to be realised through a
range of published studies including: 
a preliminary catalogue of candidate emission line objects \citep{Witham2008};
discoveries of new northern symbiotic stars \citep{Corradi2008, Corradi2010}; 
new cataclysmic variables \citep{Witham2007}; 
new groups of young stellar objects \citep{Vink2008,Barentsen2011a};
along with discoveries of new and remarkable planetary nebulae 
\citep{Mampaso2006, Corradi2011, Viironen2011}.

Over the years it has become apparent that the legacy of IPHAS 
will extend beyond the traditional \ha\ applications 
of identifying emission-line stars and nebulae. 
Through the provision of $r$, $i$ broadband photometry 
alongside H$\alpha$ data,
IPHAS has created the opportunity 
to study Galactic Plane populations 
in a new way.
For example, the survey’s unique $r-H\alpha$ colour, 
when combined with $r-i$,
has been shown to provide simultaneous constraints 
on intrinsic stellar colour and interstellar extinction \citep{Drew2008}. 
This has opened the door 
to a wide range of Galactic science applications, 
including the mapping of extinction across the Plane in three dimensions
and the probabilistic inference of stellar properties
\citep{Sale2009, Sale2010, Giammanco2011, Sale2012, Barentsen2013}. 
In effect, the availability of narrowband H$\alpha$ 
alongside $r, i$ magnitudes
provides coarse spectral information for huge samples of stars 
which are otherwise too faint or numerous 
to be targeted by spectroscopic surveys.
For such science applications to succeed however, 
it is vital that the imaging data is transformed 
into a homogeneously calibrated photometric catalogue, 
in which quality problems 
and duplicate detections are flagged.

The first release of IPHAS data, 
covering roughly half the survey footprint,
was made in late 2007 \citep{Gonzalez-Solares2008}. 
At the time the data were insufficiently complete 
to support a homogeneously calibrated source catalogue.
The goal of this paper is to present the next release 
that takes the coverage up to over 90 percent of the survey area 
and includes a uniform calibration.
In this work we aim to
(i) explain the data reduction 
and quality control procedures that were applied,
(ii) describe and test the new global photometric calibration, and 
(iii) detail the construction of the source catalogue
and demonstrate its use.

In \S\ref{sec:observations} we start by recapitulating the key points
of the survey observing strategy.
In \S\ref{sec:reduction} we describe the data reduction
and quality control procedures.
In \S\ref{sec:calibration} we explain the uniform re-calibration
and test our results against the Sloan Digital Sky Survey (SDSS).
In \S\ref{sec:catalogue} we explain how the source catalogue was compiled.
In \S\ref{sec:discussion} we discuss the properties of the catalogue,
and finally in \S\ref{sec:demonstration} we demonstrate
the scientific exploitation of the colour/magnitude diagrams
using the new catalogue.
In \S\ref{sec:conclusions} we conclude and outline
our future ambitions.

@@

The Wide Field Camera is a mosaic of 4 CCDs 
that captures a sky area of close to 0.29 square degrees.
To cover the entire Northern Plane with some overlap,
the survey area was divided into 7635 telescope pointings.
Each of these pointings is accompanied by an offset position
at a displacement of $+$5 arcmin in declination 
and $+$5 arcmin in right ascension,
to deal with inter-CCD gaps, detector imperfections,
and to enable quality checks. 
The basic unit of observation hence
amounts to $2 \times 3$ exposures, 
in which each of the 3 survey filters is exposed at 2 offset sky positions 
within an elapsed time of 10 minutes.
We shall refer to the unit of 3 exposures at the same position 
as a \emph{field},
and the combination of two fields at a small offset as a \emph{field pair}.
The survey hence contains 15\,270 fields grouped into 7\,635 field pairs.
To achieve the desired survey depth of 20th magnitude or fainter, 
the filter exposure times were set at 120 sec (narrowband \ha), 
30 sec ($r$) and 10 sec ($i$)
in the majority of the survey observations.\footnote{In 2003 
the $r$-band exposure time was 10~sec instead of 30~sec,
and since Oct 2010 the $i$-band exposure time 
has been increased from 10~sec to 20~sec.}

Data-taking began in the second half of 2003, 
and every field had been observed at least once by the end of 2008.
At that time only 76\% of the field pairs 
satisfied our minimum quality criteria however,  
often due to the effects of clouds, poor seeing, or technical faults
(the quality criteria will be detailed in the next section). 
Since then, a programme of repeat observations has been in place 
to improve data quality. 
As a result, 92\% of the survey 
now benefits from quality-approved data.
The most recent observations which are part of this release
were obtained in November 2012.

\begin{figure*}
        \includegraphics[width=1\linewidth]{./plots/footprint/footprint_small.png}
        \caption{Survey area showing the footprints
        of all the quality-approved IPHAS fields
        which have been included in this data release.
        The area covered by each field has been coloured black
        with a semi-transparent opacity of 20\%,
        such that regions where fields overlap are darker.
        The IPHAS strategy is to observe each field twice
        with a small offset,
        and hence the vast majority of the area 
        is covered twice (dominant gray colour).
        There are small overlaps between all the neighbouring fields
        which can be seen as a honeycomb-style pattern
        of dark gray lines across the survey area.
        Regions with incomplete data are apparent as white gaps (no data) 
        or in light gray (denoting that only the offset position is missing).}
        \label{fig:footprint}
\end{figure*}

Figure~\ref{fig:footprint} shows the footprint 
of the quality-approved observations included in this work. 
The fields which remain missing 
-- covering 8\% of the survey area --
are predominantly located towards the Galactic anti-center 
at $l > 120^o$.
Fields at these longitudes can only be accessed from La Palma 
in the months of November-December,
which is when the weather and seeing conditions are often poor
at the INT and observing attempts have failed repeatedly.
To enable the survey to be brought to completion, 
a decision was made recently to limit repeats in this area 
to individual fields requiring replacement,
i.e. fresh observations in all 3 filters may only be obtained 
at one of the two offset positions.  
The catalogue is structured such that it is clear 
where a contemporaneous observation of both halves of a field pair
is not available.

@@

The reduced images are then stored in multi-extension FITS files
with a primary header describing the characteristics
(position, filter, exposure time, etc.) 
and four image extensions 
corresponding to each of the four CCDs.
Source detection and characterisation is then carried out 
using the \textsc{imcore} tool \citep{Irwin1985,Irwin1997}.
The flux of each source is measured using both
the peak pixel height (i.e. a square 0.33$\times$0.33\arcsec\ aperture)
as well as a series of circular apertures of increasing diameter 
(1.2\arcsec, 2.3\arcsec, 3.3\arcsec, 4.6\arcsec\ and 6.6\arcsec).

The local background levels are estimated 
by computing the sigma-clipped median
flux in a grid of 64$\times$64 pixels (21$\times$21\arcsec)
across the image,
which is then interpolated to obtain an estimate 
of the background level at each pixel.
These sky levels are subtracted from the aperture photometry and
-- when required --
a deblending routine is applied which attempts to remove
the contamination from any nearby sources.
Whilst this approach works very well 
across the vast majority of the survey area,
the Galactic Plane unavoidably contains crowded regions 
with large numbers of overlapping sources
or rapidly spatially-varying nebulosity,
in which case aperture photometry must always be interpreted 
with caution.
In \S\ref{sec:catalogue} we will explain that objects
which suffer from this problem 
are flagged in the catalogue using the \emph{deblend} warning flag.

@@

At the time of preparing DR2,
the CASU pipeline had processed
74\,195 IPHAS exposures 
in which a total of 1.9~billion \emph{candidate sources} were detected 
at the sensitive default detection level of 1.25\,$\sigma$.
This unavoidably includes spurious detections, artefacts and
duplicate detections,
in \S\ref{sec:catalogue} we will explain
how these have been removed or flagged in the final catalogue.
The pipelined data set -- comprising 2.5~terabyte of FITS files --
was then transferred to the University of Hertfordshire
for the purpose of transforming the raw
detection tables into a reliable source catalogue which is 
(i) quality-controlled,
(ii) homogeneously calibrated, and 
(iii) contains user-friendly columns and warning flags.
It is these post-processing steps which are explained next.

@@

Observing time for IPHAS was obtained
on a semester-by-semester basis
through the traditional time allocation committees 
of the Isaac Newton Group of telescopes,
which are invariably over-subscribed.
For this reason, we attempted to utilise 
\emph{all} the nights allocated to IPHAS,
even those which were partially or entirely non-photometric
or otherwise affected by technical problems 
(e.g. electronic noise or telescope tracking problems).
Any unsuitable data that was taken as a result
of this strategy has subsequently been flagged and removed
using a series of seven quality criteria,
which ensure a reliable and homogeneous level of quality
across the data release:

@@

(1) \emph{Depth.} 
We discarded any exposures for which the $5\sigma$ limiting magnitude 
was worse than 20th magnitude in the $r$-band
or worse than 19th in $i$ or \ha. 
Such data were typically obtained during poor weather or full moon.
Most observations fared significantly better than these limits.
Figure~\ref{fig:depth} presents the distribution of limiting magnitudes
for all quality-approved fields,
showing a mean depth of 
$21.2\pm0.5$ ($r$), $20.0\pm0.3$ ($i$) and $20.3\pm0.3$ (\ha).
The depth achieved depended 
most strongly on the presence of the moon,
which was above the horizon during 62\% 
of the observations and explains the wide and bi-modal shape
of the $r$-band limiting magnitude distribution 
(top panel in Fig.~\ref{fig:depth}).
In contrast, the depth attained in $i$ and \ha\ 
is less sensitive to moonlight
and their magnitude limit distributions
are hence more narrow
(middle and bottom panels in Fig.~\ref{fig:depth}).

The minimum depth criterion
has led us to exclude 9\% of the pipelined data,
which were often obtained under cloudy conditions.
We note that some of the excluded data may nevertheless be useful
for e.g. time-domain studies of bright stars.
The detection tables of the discarded data are made
available through our website for this reason (www.iphas.org),
but are ignored in what follows.

(2) \emph{Ellipticity.} 
The ellipticity of a point source,
defined as $e = 1 - b / a$ 
with $b$ the semi-minor and $a$ the semi-major axis,
is a morphological measure of the elongation of the point spread function.
It is expected to be zero (circular) across the field 
in a perfect imaging system,
but it is slightly non-zero in any real telescope data 
due to optical distortions and tracking errors.
The mean ellipticity across a field in the IPHAS data set 
is $0.09\pm0.04$.
There have been sporadic episodes with higher ellipticities
due to mechanical glitches in the telescope tracking system.
For this reason, we rejected exposures in which the mean ellipticity
across the detector exceeded $e > 0.3$,
which is when the photometric measurements delivered by the pipeline
were found to become degraded.
Only 0.4\% of the exposures were discarded on this basis.

(3) \emph{Seeing.} 
The survey originally aimed to obtain data 
at seeing better than 1.7 arcsec.
This target is currently attained across 86\% of the footprint,
in particular at early longitudes,
e.g. 92\% of the fields at $l<120^\circ$ are better than 1.7 arcsec.
Figure~\ref{fig:seeing} presents the distribution
of the mean seeing for all the quality-approved fields.
We find a median value of 1.1~arcsec in $r$/\ha\
and 1.0~arcsec in $i$.
In the $r$-band, 90\% of the data is better than 1.5~arcsec
and 10\% is better than 0.8~arcsec.
To improve the area covered by this data release,
we have decided to include the small fraction of data
that was obtained under seeing up to 2.5 arcsec,
so that only 1\%of the pipelined exposures
had to be excluded due to poor seeing.
In \S\ref{sec:catalogue} we will explain
that the photometry listed in the catalogue
is predominantly based on the images with the
best-available seeing, and that the seeing estimate
is included as a column in the catalogue.
